---
title: "Can We Predict Abalone Age Without Cutting It Open?"
author: "Beakale Amenu Tefera"
date: "2026"
output: html_document
---
## 1: Introduction & Data Loading

In this project, we ask:
**Can we predict the age of an abalone from its physical 
measurements, without cutting it open?**

An abalone is a sea snail. A marine shellfish found in cold coastal
waters around the world. It is considered a delicacy in many countries, 
especially in East Asia, and is commercially farmed and harvested.

Normally, age is determined by cutting the shell and counting 
rings under a microscope. This project explores whether machine 
learning can replace that process using only external measurements.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-data}
# Load libraries
library(ggplot2)
library(dplyr)

# Load the dataset
abalone <- read.csv("/home/beakale/Desktop/Projects/abalone_project/abalone.data", 
                    header = FALSE)

# Assign column names
colnames(abalone) <- c("Sex", "Length", "Diameter", "Height",
                       "Whole_weight", "Shucked_weight", 
                       "Viscera_weight", "Shell_weight", "Rings")

# Fix Sex as a factor
abalone$Sex <- as.factor(abalone$Sex)

# Age = Rings + 1.5
abalone$Age <- abalone$Rings + 1.5

# Check structure
str(abalone)
```
## 2: Visualising the Data

Before building any model, we first need to understand 
what we are trying to predict. How is Age distributed? 
Is it balanced or skewed?

```{r age-distribution}
ggplot(abalone, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "#2E6096", color = "white") +
  labs(
    title = "Distribution of Abalone Age",
    subtitle = "Right skewed with peak at age 10, tail stretching to 24.5 years",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r boxplot-sex}
ggplot(abalone, aes(x=Sex, y=Age, fill=Sex)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=8, outlier.size=4) +
  scale_fill_manual(values=c(
    "F" = "pink",
    "I" = "lavender",
    "M" = "skyblue"
  )) +
  labs(
    title="Abalone Age by Sex",
    subtitle="Males show the highest median age, females the most outliers, and infants the youngest distribution",
    x="Sex",
    y="Age (years)"
  ) +
  theme_minimal()
```
```{r pairwise-plot}
library(GGally)

ggpairs(abalone, 
        columns = c("Length", "Diameter", "Height",
                    "Whole_weight", "Shucked_weight",
                    "Viscera_weight", "Shell_weight"),
        aes(color = Sex, alpha = 0.5)) +
  labs(
    title = "Pairwise Plot of Physical Measurements",
    subtitle = "Male and Female (Left skewed) and Infant (Symmetric), all are highly correlated"
  ) +
  theme_minimal()
```
**Observation:** Male and Female distributions are left skewed 
while Infant is symmetric. All physical measurements are highly 
correlated, indicating multicollinearity.
```{r descriptive-stats}
library(psych)

# Descriptive statistics for all numeric columns
describe(abalone[, c("Length", "Diameter", "Height",
                      "Whole_weight", "Shucked_weight",
                      "Viscera_weight", "Shell_weight", "Age")])
```
The descriptive statistics confirm the visual observations. Length and Diameter are negatively skewed reflecting that most abalones are fully grown, while Height is strongly positively skewed (3.13) indicating most abalones remain flat with few growing tall. Variables with similar mean and median values suggest symmetric distributions. Age shows the largest standard deviation, suggesting that abalones of similar size can vary considerably in age, highlighting the challenge of accurate age prediction.

H0 — "There is no difference in Age between Male, Female and Infant abalones"
H1 — "There is a significant difference in Age between at least one group"
```{r hypothesis-testing}
# Kruskal-Wallis test - comparing Age across Sex groups
result <- kruskal.test(Age ~ Sex, data = abalone)

# Clean output
cat("Kruskal-Wallis Test: Age by Sex\n")
cat("================================\n")
cat("Chi-squared:", round(result$statistic, 3), "\n")
cat("Degrees of freedom:", result$parameter, "\n")
cat("P-value:", format(result$p.value, scientific = TRUE), "\n")
cat("================================\n")
cat("Conclusion: We reject H0 - significant difference in Age between groups\n")
```
Based on the Kruskal–Wallis test, Sex has a statistically significant effect on Age.
```{r correlation}
# Correlation of all numeric variables with Age
cor(abalone[, c("Length", "Diameter", "Height",
                 "Whole_weight", "Shucked_weight",
                 "Viscera_weight", "Shell_weight", "Age")])
```
Shucked weight is the lowest because it represents the soft body meat which varies based on factors other than age such as nutrition and season.
```{r correlation-heatmap}
library(ggplot2)
library(reshape2)

# Correlation matrix
corr_matrix <- cor(abalone[, sapply(abalone, is.numeric)], use = "complete.obs")

# Convert to long format
corr_long <- melt(corr_matrix)

# Heatmap
ggplot(corr_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limit = c(-1, 1), name = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    axis.title = element_blank()
  ) +
  coord_fixed() +
  labs(
    title = "Correlation Heatmap",
    subtitle = "Relationships Between Numerical Variables"
  )
```
Shell weight is the strongest predictor of Age, while even the weakest correlation (0.42) suggests all physical measurements carry some predictive value.
```{r simple-regression}
# Simple Linear Regression - Age predicted by Shell weight
simple_model <- lm(Age ~ Shell_weight, data = abalone)
summary(simple_model)
```
```{r simple-regression-plot}
ggplot(abalone, aes(x=Shell_weight, y=Age)) +
  geom_point(alpha=0.3, color="#2E6096") +
  geom_smooth(method="lm", color="red", se=TRUE) +
  labs(
    title="Simple Linear Regression: Age ~ Shell Weight",
    x="Shell Weight",
    y="Age (years)"
  ) +
  theme_minimal()
```
he simple linear regression model using Shell Weight as the sole predictor produced an R² of 0.39, meaning the model explains only 39% of the variation in Age. The scatter plot revealed a non-linear pattern, suggesting the relationship is significant but not perfectly linear. This indicates that one variable is not sufficient, and motivates the use of Multiple Linear Regression.
## 6.2 Multiple Linear Regression — All Predictors

We now build a model using all physical measurements plus Sex 
as predictors. The goal is to test whether combining variables 
improves prediction over Shell Weight alone.

```{r mlr-model}
# Build the full multiple linear regression model
# We list every predictor explicitly so it is clear what goes in

mlr_model <- lm(Age ~ Shell_weight + Length + Diameter + Height +
                  Whole_weight + Shucked_weight + Viscera_weight + Sex,
                data = abalone)

# Print the summary — R², p-values, and coefficients
summary(mlr_model)
```
## 6.2 Conclusion — Multiple Linear Regression

The multiple linear regression model produced an Adjusted R² of 0.54, 
explaining 54% of the variation in Age — an improvement over the simple 
model. However, **Length** and **Sex (Male)** were not statistically 
significant. The negative coefficient on Shucked Weight suggests 
multicollinearity is affecting the model, consistent with the 
correlations observed in Section 5.
## 6.3 Model Diagnostics

Before trusting any regression model, we check four diagnostic 
plots. These tell us whether the assumptions of linear regression 
are being met.

```{r mlr-diagnostics}
# Set 2x2 layout
par(mfrow = c(2, 2))

# Plot 1 — Residuals vs Fitted
plot(mlr_model, which = 1,
     col = "#E74C3C",
     pch = 19,
     main = "Residuals vs Fitted")

# Plot 2 — Q-Q Plot
plot(mlr_model, which = 2,
     col = "#2ECC71",
     pch = 19,
     main = "Normal Q-Q")

# Plot 3 — Scale-Location
plot(mlr_model, which = 3,
     col = "#3498DB",
     pch = 19,
     main = "Scale-Location")

# Plot 4 — Residuals vs Leverage
plot(mlr_model, which = 4,
     col = "#9B59B6",
     pch = 19,
     main = "Residuals vs Leverage")

# Reset layout
par(mfrow = c(1, 1))
```
## 6.3 Conclusion — Model Diagnostics

The diagnostic plots reveal that the residuals show a non-linear 
pattern, confirming the relationship is not perfectly linear. 
The Q-Q plot suggests residuals are approximately normal, with 
slight deviation at the tails. The Scale-Location plot indicates 
heteroscedasticity, meaning the model makes larger errors for some 
predicted values. Three influential observations were identified 
in the Residuals vs Leverage plot, which may be affecting the 
model coefficients.
## 6.4 Predicted vs Actual Age

We now compare what the model predicted against the actual 
known ages. If the model were perfect, every point would 
fall exactly on the diagonal line.

```{r predicted-vs-actual}
# Add predictions from our MLR model back into the dataset
abalone$Predicted_Age <- predict(mlr_model, abalone)

# Plot Predicted vs Actual
ggplot(abalone, aes(x = Age, y = Predicted_Age)) +
  geom_point(alpha = 0.3, color = "#3498DB") +
  geom_abline(intercept = 0, slope = 1,
              color = "#E74C3C",
              linewidth = 1,
              linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Age — Multiple Linear Regression",
    x = "Actual Age",
    y = "Predicted Age"
  ) +
  theme_minimal()
```
## 6.4 Conclusion — Predicted vs Actual Age

The Predicted vs Actual plot shows the model performs best for 
middle aged abalones. For young abalones the model tends to 
overestimate, while for older abalones it tends to underestimate. 
This pattern is consistent with the non-linear growth curve 
observed in Section 6.1, where abalone growth slows as the 
animal ages.

## 6.5 Section 6 Summary — Linear Regression

**Simple Linear Regression**
- Predictor: Shell Weight only
- R²: 0.39 — explains 39% of variation in Age
- Limitation: non-linear relationship, one variable not enough

**Multiple Linear Regression**
- Predictors: All physical measurements + Sex
- Adjusted R²: 0.54 — explains 54% of variation in Age
- Non-significant: Length, Sex (Male)
- Limitations: non-linear residuals, heteroscedasticity, 
  three influential observations

**Overall Conclusion**

Linear regression established that physical measurements carry 
meaningful predictive signal for abalone age. However, the 
non-linear growth pattern and multicollinearity among predictors 
limit the model. These findings motivate the use of classification 
methods and more robust models in the sections that follow.


## 7.1 Age Group Classification

Rather than predicting exact age, we now classify each abalone 
into one of three groups — Young, Middle, or Old — based on 
the age distribution observed in Section 2.

## 7.1 Age Group Classification

Rather than predicting exact age, we now classify each abalone 
into one of three groups — Young, Middle, or Old — based on 
the age distribution observed in Section 2.

```{r age-groups}
# Cut Age into three groups using our defined boundaries
abalone$Age_Group <- cut(abalone$Age,
                         breaks = c(0, 9, 13, Inf),
                         labels = c("Young", "Middle", "Old"))

# Check counts and percentages
age_counts <- table(abalone$Age_Group)
age_pct <- round(prop.table(age_counts) * 100, 1)
age_counts
age_pct
```
## 7.2 Age Group Distribution Plot

```{r age-group-plot}
ggplot(abalone, aes(x = Age, fill = Age_Group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Young"  = "#3498DB",
                               "Middle" = "#2ECC71",
                               "Old"    = "#95A5A6")) +
  geom_vline(xintercept = c(9, 13),
             linetype = "dashed",
             color = "black",
             linewidth = 0.7) +
  labs(
    title = "Density Distribution of Abalone Age Groups",
    subtitle = "",
    x = "Age",
    y = "Density",
    fill = "Age Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```
## 7.2 Conclusion — Age Group Distribution

The density plot reveals that Young abalones peak near the Age 9 
boundary, suggesting most young abalones are clustered at upper 
ages. Middle aged abalones show a uniform distribution, confirming 
they are the dominant group. Old abalones have a low flat density, 
consistent with the right skew observed in Section 2.
## 7.3 Random Forest Classification

Random Forest builds multiple decision trees and combines their 
predictions by majority vote. It handles multicollinearity better 
than linear regression and provides feature importance scores 
showing which variables contributed most to classification.

```{r random-forest}
# Load the library
library(randomForest)

# Set seed so results are reproducible
set.seed(42)

# Train the Random Forest model
rf_model <- randomForest(Age_Group ~ Shell_weight + Length + 
                           Diameter + Height + Whole_weight + 
                           Shucked_weight + Viscera_weight + Sex,
                         data = abalone,
                         ntree = 500,
                         importance = TRUE)
```

```{r rf-output}
# Clean presentation of Random Forest results
cat("Random Forest Model Summary\n")
cat("---------------------------\n")
cat("Number of Trees     :", rf_model$ntree, "\n")
cat("Number of Predictors:", rf_model$mtry, "\n")
cat("---------------------------\n")
cat("Overall OOB Error   : 26.31%\n")
cat("---------------------------\n")
cat("Error by Age Group\n")
cat("---------------------------\n")
cat("Young  :", round(rf_model$confusion["Young",  "class.error"] * 100, 2), "%\n")
cat("Middle :", round(rf_model$confusion["Middle", "class.error"] * 100, 2), "%\n")
cat("Old    :", round(rf_model$confusion["Old",    "class.error"] * 100, 2), "%\n")
cat("---------------------------\n")
cat("Overall Accuracy    :", round((1 - 0.2631) * 100, 2), "%\n")
```
```{r feature-importance}
# Extract importance scores from the model
importance_df <- data.frame(
  Variable = rownames(importance(rf_model)),
  Importance = importance(rf_model)[, "MeanDecreaseGini"]
)

# Order by importance
importance_df <- importance_df[order(importance_df$Importance,
                                     decreasing = FALSE), ]

# Lollipop Chart
ggplot(importance_df, aes(x = reorder(Variable, Importance),
                          y = Importance)) +
  geom_segment(aes(x = reorder(Variable, Importance),
                   xend = reorder(Variable, Importance),
                   y = 0,
                   yend = Importance),
               color = "#BDC3C7",
               linewidth = 0.8) +
  geom_point(aes(color = Importance),
             size = 5) +
  scale_color_gradient(low = "#3498DB",
                       high = "#E74C3C") +
  coord_flip() +
  labs(
    title = "Feature Importance — Random Forest",
    subtitle = "",
    x = "Variable",
    y = "Mean Decrease in Gini"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none",
    panel.grid.major.y = element_blank()
  )
```
## 7.4 Conclusion — Feature Importance

The feature importance plot reveals that Shell Weight is the 
strongest predictor of Age Group, consistent with the correlation 
heatmap in Section 5. Sex has the lowest importance score, 
suggesting that while Sex has a statistically significant effect 
on Age as confirmed in Section 4, physical measurements carry 
greater predictive power. The remaining variables contribute 
similar and moderate importance, reflecting the multicollinearity 
observed throughout the analysis.

## 7.5 Section 7 Summary — Classification

**Age Group Definition**
- Young  : Age ≤ 9
- Middle : 9 < Age ≤ 13
- Old    : Age > 13

**Class Distribution**
- Middle is the dominant group — class imbalance confirmed
- Young abalones peak near the Age 9 boundary
- Old abalones are sparse — consistent with right skew

**Random Forest Classification**
- Overall Accuracy : 73.69%
- Overall OOB Error: 26.31%
- Best classified   : Middle — lowest error
- Worst classified  : Old — highest error
- Reason            : class imbalance, Old is underrepresented

**Feature Importance**
- Strongest predictor : Shell Weight
- Weakest predictor   : Sex
- Remaining variables : similar and moderate importance

**Overall Conclusion**

Random Forest achieved 73.69% accuracy in classifying abalone 
age groups. However, class imbalance causes the model to 
underperform on Old abalones. Shell Weight remains the strongest 
predictor across both regression and classification methods. 
These findings motivate the use of K-Fold Cross Validation in 
Section 8 to confirm model reliability, and SMOTE in Section 9 
to address class imbalance.
## 8.1 K-Fold Cross Validation

K-Fold Cross Validation tests whether the model accuracy is 
reliable or inflated by overfitting. The data is split into 
10 equal folds. The model trains on 9 folds and tests on 1, 
repeating 10 times. The final accuracy is the average across 
all 10 folds.

```{r kfold-output}
# Extract the best accuracy from cross validation results
cv_accuracy <- max(rf_cv$results$Accuracy) * 100
cv_kappa <- max(rf_cv$results$Kappa)

cat("K-Fold Cross Validation Results\n")
cat("--------------------------------\n")
cat("Method          : 10-Fold Cross Validation\n")
cat("Model           : Random Forest\n")
cat("Number of Trees : 500\n")
cat("--------------------------------\n")
cat("CV Accuracy     :", round(cv_accuracy, 2), "%\n")
cat("Kappa Score     :", round(cv_kappa, 3), "\n")
cat("--------------------------------\n")
cat("Original OOB    : 73.69%\n")
cat("Difference      :", round(cv_accuracy - 73.69, 2), "%\n")
cat("--------------------------------\n")
```
## 8.1 Conclusion — K-Fold Cross Validation

K-Fold Cross Validation confirmed that the Random Forest accuracy 
of 73.69% is reliable, with cross validated accuracy of 73.52% — 
a difference of only 0.17%. The Kappa score of 0.524 indicates 
moderate agreement, suggesting the model performs meaningfully 
better than chance. However, Old abalones show a classification 
error of 51.88%, confirming that class imbalance remains the 
primary limitation to address in Section 9.

## 8.2 Section 8 Summary — K-Fold Cross Validation

**Cross Validation Setup**
- Method  : 10-Fold Cross Validation
- Model   : Random Forest
- Trees   : 500

**Results**
- Original Accuracy : 73.69%
- CV Accuracy       : 73.52%
- Difference        : -0.17%
- Kappa Score       : 0.524 — moderate agreement

**Key Finding**
- Difference of 0.17% confirms no overfitting
- Model genuinely learned patterns in the data
- Kappa of 0.524 confirms performance above chance

**Remaining Limitation**
- Old abalone error : 51.88%
- Cause             : class imbalance
- Middle dominates  : 56.9% of all observations

**Overall Conclusion**

Cross validation confirmed the Random Forest model is reliable 
and not overfitting. However, the 51.88% error rate on Old 
abalones reveals that class imbalance is significantly affecting 
model performance. Section 9 addresses this directly using SMOTE 
to balance the classes before retraining.
## 9.1 SMOTE — Balancing the Classes

The Random Forest model achieved only 48.12% accuracy on Old 
abalones due to class imbalance. SMOTE creates synthetic 
observations for the minority classes — Young and Old — until 
all three groups are balanced. We then retrain the Random Forest 
on the balanced dataset and compare performance.

```{r smote}
library(smotefamily)

# model.matrix converts Sex to dummy variables (0/1)
abalone_numeric <- data.frame(
  model.matrix(~ Sex - 1, data = abalone),
  Shell_weight   = abalone$Shell_weight,
  Length         = abalone$Length,
  Diameter       = abalone$Diameter,
  Height         = abalone$Height,
  Whole_weight   = abalone$Whole_weight,
  Shucked_weight = abalone$Shucked_weight,
  Viscera_weight = abalone$Viscera_weight
)

# Keep Age_Group separate as the target label
age_group_label <- as.character(abalone$Age_Group)

# Split by class
young_idx  <- age_group_label == "Young"
middle_idx <- age_group_label == "Middle"
old_idx    <- age_group_label == "Old"

young_data  <- abalone_numeric[young_idx,  ]
middle_data <- abalone_numeric[middle_idx, ]
old_data    <- abalone_numeric[old_idx,    ]

young_label  <- age_group_label[young_idx]
old_label    <- age_group_label[old_idx]

# Target size matches Middle
target <- nrow(middle_data)

# SMOTE Young up to Middle size
set.seed(42)
smote_old <- SMOTE(old_data,
                     old_label,
                     K = 5,
                     dup_size = ceiling(target / nrow(old_data)) - 1)

# SMOTE Old up to Middle size
set.seed(42)
smote_young <- SMOTE(young_data,
                   young_label,
                   K = 5,
                   dup_size = ceiling(target / nrow(young_data)) - 1)

# Combine all three classes
abalone_smote <- rbind(
  smote_young$data,
  cbind(middle_data, class = "Middle"),
  smote_old$data
)

# Set class as factor
abalone_smote$class <- as.factor(abalone_smote$class)

# Check distribution
smote_counts <- table(abalone_smote$class)
smote_pct    <- round(prop.table(smote_counts) * 100, 1)

cat("Class Distribution After SMOTE\n")
cat("--------------------------------\n")
cat("Young  :", smote_counts["Young"],
    "—", smote_pct["Young"],  "%\n")
cat("Middle :", smote_counts["Middle"],
    "—", smote_pct["Middle"], "%\n")
cat("Old    :", smote_counts["Old"],
    "—", smote_pct["Old"],    "%\n")
cat("--------------------------------\n")
cat("Total  :", sum(smote_counts), "observations\n")
```
## 9.1 Conclusion — SMOTE

After applying SMOTE, the class distribution improved 
significantly. Young increased from 839 to 2517, Middle 
remained at 2378, and Old increased from 960 to 2880. 
The dataset now contains 7775 total observations, with 
all three classes in the 30.6 to 37% range. This balanced 
distribution allows the Random Forest to learn more 
effectively from Young and Old abalones.