---
title: "Can We Predict Abalone Age Without Cutting It Open?"
author: "Beakale Amenu Tefera"
date: "2026"
output: html_document
---
## 1: Introduction & Data Loading

In this project, we ask:
**Can we predict the age of an abalone from its physical 
measurements, without cutting it open?**

An abalone is a sea snail. A marine shellfish found in cold coastal
waters around the world. It is considered a delicacy in many countries, 
especially in East Asia, and is commercially farmed and harvested.

Normally, age is determined by cutting the shell and counting 
rings under a microscope. This project explores whether machine 
learning can replace that process using only external measurements.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-data}
# Load libraries
library(ggplot2)
library(dplyr)

# Load the dataset
abalone <- read.csv("/home/beakale/Desktop/Projects/abalone_project/abalone.data", 
                    header = FALSE)

# Assign column names
colnames(abalone) <- c("Sex", "Length", "Diameter", "Height",
                       "Whole_weight", "Shucked_weight", 
                       "Viscera_weight", "Shell_weight", "Rings")

# Fix Sex as a factor
abalone$Sex <- as.factor(abalone$Sex)

# Age = Rings + 1.5
abalone$Age <- abalone$Rings + 1.5

# Check structure
str(abalone)
```
## 2: Visualising the Data

Before building any model, we first need to understand 
what we are trying to predict. How is Age distributed? 
Is it balanced or skewed?

```{r age-distribution}
ggplot(abalone, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "#2E6096", color = "white") +
  labs(
    title = "Distribution of Abalone Age",
    subtitle = "Right skewed with peak at age 10, tail stretching to 24.5 years",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r boxplot-sex}
ggplot(abalone, aes(x=Sex, y=Age, fill=Sex)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=8, outlier.size=4) +
  scale_fill_manual(values=c(
    "F" = "pink",
    "I" = "lavender",
    "M" = "skyblue"
  )) +
  labs(
    title="Abalone Age by Sex",
    subtitle="Males show the highest median age, females the most outliers, and infants the youngest distribution",
    x="Sex",
    y="Age (years)"
  ) +
  theme_minimal()
```
```{r pairwise-plot}
library(GGally)

ggpairs(abalone, 
        columns = c("Length", "Diameter", "Height",
                    "Whole_weight", "Shucked_weight",
                    "Viscera_weight", "Shell_weight"),
        aes(color = Sex, alpha = 0.5)) +
  theme_minimal()
```