---
title: "Can We Predict Abalone Age Without Cutting It Open?"
author: "Beakale Amenu Tefera"
date: "2026"
output: html_document
---
## 1: Introduction & Data Loading

In this project, we ask:
**Can we predict the age of an abalone from its physical 
measurements, without cutting it open?**

An abalone is a sea snail. A marine shellfish found in cold coastal
waters around the world. It is considered a delicacy in many countries, 
especially in East Asia, and is commercially farmed and harvested.

Normally, age is determined by cutting the shell and counting 
rings under a microscope. This project explores whether machine 
learning can replace that process using only external measurements.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-data}
# Load libraries
library(ggplot2)
library(dplyr)

# Load the dataset
abalone <- read.csv("/home/beakale/Desktop/Projects/abalone_project/abalone.data", 
                    header = FALSE)

# Assign column names
colnames(abalone) <- c("Sex", "Length", "Diameter", "Height",
                       "Whole_weight", "Shucked_weight", 
                       "Viscera_weight", "Shell_weight", "Rings")

# Fix Sex as a factor
abalone$Sex <- as.factor(abalone$Sex)

# Age = Rings + 1.5
abalone$Age <- abalone$Rings + 1.5

# Check structure
str(abalone)
```
## 2: Visualising the Data

Before building any model, we first need to understand 
what we are trying to predict. How is Age distributed? 
Is it balanced or skewed?

```{r age-distribution}
ggplot(abalone, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "#2E6096", color = "white") +
  labs(
    title = "Distribution of Abalone Age",
    subtitle = "Right skewed with peak at age 10, tail stretching to 24.5 years",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r boxplot-sex}
ggplot(abalone, aes(x=Sex, y=Age, fill=Sex)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=8, outlier.size=4) +
  scale_fill_manual(values=c(
    "F" = "pink",
    "I" = "lavender",
    "M" = "skyblue"
  )) +
  labs(
    title="Abalone Age by Sex",
    subtitle="Males show the highest median age, females the most outliers, and infants the youngest distribution",
    x="Sex",
    y="Age (years)"
  ) +
  theme_minimal()
```
```{r pairwise-plot}
library(GGally)

ggpairs(abalone, 
        columns = c("Length", "Diameter", "Height",
                    "Whole_weight", "Shucked_weight",
                    "Viscera_weight", "Shell_weight"),
        aes(color = Sex, alpha = 0.5)) +
  labs(
    title = "Pairwise Plot of Physical Measurements",
    subtitle = "Male and Female (Left skewed) and Infant (Symmetric), all are highly correlated"
  ) +
  theme_minimal()
```
**Observation:** Male and Female distributions are left skewed 
while Infant is symmetric. All physical measurements are highly 
correlated, indicating multicollinearity.
```{r descriptive-stats}
library(psych)

# Descriptive statistics for all numeric columns
describe(abalone[, c("Length", "Diameter", "Height",
                      "Whole_weight", "Shucked_weight",
                      "Viscera_weight", "Shell_weight", "Age")])
```
The descriptive statistics confirm the visual observations. Length and Diameter are negatively skewed reflecting that most abalones are fully grown, while Height is strongly positively skewed (3.13) indicating most abalones remain flat with few growing tall. Variables with similar mean and median values suggest symmetric distributions. Age shows the largest standard deviation, suggesting that abalones of similar size can vary considerably in age, highlighting the challenge of accurate age prediction.

H0 — "There is no difference in Age between Male, Female and Infant abalones"
H1 — "There is a significant difference in Age between at least one group"
```{r hypothesis-testing}
# Kruskal-Wallis test - comparing Age across Sex groups
result <- kruskal.test(Age ~ Sex, data = abalone)

# Clean output
cat("Kruskal-Wallis Test: Age by Sex\n")
cat("================================\n")
cat("Chi-squared:", round(result$statistic, 3), "\n")
cat("Degrees of freedom:", result$parameter, "\n")
cat("P-value:", format(result$p.value, scientific = TRUE), "\n")
cat("================================\n")
cat("Conclusion: We reject H0 - significant difference in Age between groups\n")
```
Based on the Kruskal–Wallis test, Sex has a statistically significant effect on Age.
```{r correlation}
# Correlation of all numeric variables with Age
cor(abalone[, c("Length", "Diameter", "Height",
                 "Whole_weight", "Shucked_weight",
                 "Viscera_weight", "Shell_weight", "Age")])
```
Shucked weight is the lowest because it represents the soft body meat which varies based on factors other than age such as nutrition and season.
```{r correlation-heatmap}
library(ggplot2)
library(reshape2)

# Correlation matrix
corr_matrix <- cor(abalone[, sapply(abalone, is.numeric)], use = "complete.obs")

# Convert to long format
corr_long <- melt(corr_matrix)

# Heatmap
ggplot(corr_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limit = c(-1, 1), name = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    axis.title = element_blank()
  ) +
  coord_fixed() +
  labs(
    title = "Correlation Heatmap",
    subtitle = "Relationships Between Numerical Variables"
  )
```
Shell weight is the strongest predictor of Age, while even the weakest correlation (0.42) suggests all physical measurements carry some predictive value.
```{r simple-regression}
# Simple Linear Regression - Age predicted by Shell weight
simple_model <- lm(Age ~ Shell_weight, data = abalone)
summary(simple_model)
```
```{r simple-regression-plot}
ggplot(abalone, aes(x=Shell_weight, y=Age)) +
  geom_point(alpha=0.3, color="#2E6096") +
  geom_smooth(method="lm", color="red", se=TRUE) +
  labs(
    title="Simple Linear Regression: Age ~ Shell Weight",
    x="Shell Weight",
    y="Age (years)"
  ) +
  theme_minimal()
```
he simple linear regression model using Shell Weight as the sole predictor produced an R² of 0.39, meaning the model explains only 39% of the variation in Age. The scatter plot revealed a non-linear pattern, suggesting the relationship is significant but not perfectly linear. This indicates that one variable is not sufficient, and motivates the use of Multiple Linear Regression.
## 6.2 Multiple Linear Regression — All Predictors

We now build a model using all physical measurements plus Sex 
as predictors. The goal is to test whether combining variables 
improves prediction over Shell Weight alone.

```{r mlr-model}
# Build the full multiple linear regression model
# We list every predictor explicitly so it is clear what goes in

mlr_model <- lm(Age ~ Shell_weight + Length + Diameter + Height +
                  Whole_weight + Shucked_weight + Viscera_weight + Sex,
                data = abalone)

# Print the summary — R², p-values, and coefficients
summary(mlr_model)
```
## 6.2 Conclusion — Multiple Linear Regression

The multiple linear regression model produced an Adjusted R² of 0.54, 
explaining 54% of the variation in Age — an improvement over the simple 
model. However, **Length** and **Sex (Male)** were not statistically 
significant. The negative coefficient on Shucked Weight suggests 
multicollinearity is affecting the model, consistent with the 
correlations observed in Section 5.
## 6.3 Model Diagnostics

Before trusting any regression model, we check four diagnostic 
plots. These tell us whether the assumptions of linear regression 
are being met.

```{r mlr-diagnostics}
# Set 2x2 layout
par(mfrow = c(2, 2))

# Plot 1 — Residuals vs Fitted
plot(mlr_model, which = 1,
     col = "#E74C3C",
     pch = 19,
     main = "Residuals vs Fitted")

# Plot 2 — Q-Q Plot
plot(mlr_model, which = 2,
     col = "#2ECC71",
     pch = 19,
     main = "Normal Q-Q")

# Plot 3 — Scale-Location
plot(mlr_model, which = 3,
     col = "#3498DB",
     pch = 19,
     main = "Scale-Location")

# Plot 4 — Residuals vs Leverage
plot(mlr_model, which = 4,
     col = "#9B59B6",
     pch = 19,
     main = "Residuals vs Leverage")

# Reset layout
par(mfrow = c(1, 1))
```
## 6.3 Conclusion — Model Diagnostics

The diagnostic plots reveal that the residuals show a non-linear 
pattern, confirming the relationship is not perfectly linear. 
The Q-Q plot suggests residuals are approximately normal, with 
slight deviation at the tails. The Scale-Location plot indicates 
heteroscedasticity, meaning the model makes larger errors for some 
predicted values. Three influential observations were identified 
in the Residuals vs Leverage plot, which may be affecting the 
model coefficients.
## 6.4 Predicted vs Actual Age

We now compare what the model predicted against the actual 
known ages. If the model were perfect, every point would 
fall exactly on the diagonal line.

```{r predicted-vs-actual}
# Add predictions from our MLR model back into the dataset
abalone$Predicted_Age <- predict(mlr_model, abalone)

# Plot Predicted vs Actual
ggplot(abalone, aes(x = Age, y = Predicted_Age)) +
  geom_point(alpha = 0.3, color = "#3498DB") +
  geom_abline(intercept = 0, slope = 1,
              color = "#E74C3C",
              linewidth = 1,
              linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Age — Multiple Linear Regression",
    x = "Actual Age",
    y = "Predicted Age"
  ) +
  theme_minimal()
```
## 6.4 Conclusion — Predicted vs Actual Age

The Predicted vs Actual plot shows the model performs best for 
middle aged abalones. For young abalones the model tends to 
overestimate, while for older abalones it tends to underestimate. 
This pattern is consistent with the non-linear growth curve 
observed in Section 6.1, where abalone growth slows as the 
animal ages.
library(knitr)
library(kableExtra)

# Cleaner comparison table
model_comparison <- data.frame(
  Model = c(
    "Simple Linear Regression",
    "Multiple Linear Regression"
  ),

  Predictors = c(
    "Shell Weight",
    "Physical Measurements + Sex"
  ),

  `R²` = c(0.39, 0.54),

  `Adjusted R²` = c(0.38, 0.54),

  `Key Significant Predictors` = c(
    "Shell Weight",
    "Shell Weight, Diameter,\nHeight, Whole Weight,\nShucked Weight,\nViscera Weight,\nSex (Infant)"
  ),

  Main_Limitation = c(
    "Non-linear relationship",
    "Multicollinearity &\nheteroscedasticity"
  )
)

# Professional table
kable(
  model_comparison,
  caption = "Section 6 Summary — Linear Regression Models",
  align = "c",
  booktabs = TRUE
) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed",
      "responsive"
    ),
    font_size = 13,
    position = "center"
  ) %>%
  row_spec(0,
           bold = TRUE,
           color = "white",
           background = "#2C3E50")

Linear regression established that physical measurements carry 
meaningful predictive signal for abalone age. Shell Weight alone 
explained 39% of the variation in Age. Adding all physical 
measurements and Sex improved this to 54%. However, diagnostic 
plots confirmed three limitations — a non-linear residual pattern, 
heteroscedasticity, and three influential observations. The model 
performs best for middle aged abalones, overpredicting for young 
and underpredicting for old. These limitations motivate the use 
of classification methods and more robust models in the sections 
that follow.