---
title: "Can We Predict Abalone Age Without Cutting It Open?"
author: "Beakale Amenu Tefera"
date: "2026"
output: html_document
---
## 1: Introduction & Data Loading

In this project, we ask:
**Can we predict the age of an abalone from its physical 
measurements, without cutting it open?**

An abalone is a sea snail. A marine shellfish found in cold coastal
waters around the world. It is considered a delicacy in many countries, 
especially in East Asia, and is commercially farmed and harvested.

Normally, age is determined by cutting the shell and counting 
rings under a microscope. This project explores whether machine 
learning can replace that process using only external measurements.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-data}
# Load libraries
library(ggplot2)
library(dplyr)

# Load the dataset
abalone <- read.csv("/home/beakale/Desktop/Projects/abalone_project/abalone.data", 
                    header = FALSE)

# Assign column names
colnames(abalone) <- c("Sex", "Length", "Diameter", "Height",
                       "Whole_weight", "Shucked_weight", 
                       "Viscera_weight", "Shell_weight", "Rings")

# Fix Sex as a factor
abalone$Sex <- as.factor(abalone$Sex)

# Age = Rings + 1.5
abalone$Age <- abalone$Rings + 1.5

# Check structure
str(abalone)
```
## 2: Visualising the Data

Before building any model, we first need to understand 
what we are trying to predict. How is Age distributed? 
Is it balanced or skewed?

```{r age-distribution}
ggplot(abalone, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "#2E6096", color = "white") +
  labs(
    title = "Distribution of Abalone Age",
    subtitle = "Right skewed with peak at age 10, tail stretching to 24.5 years",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r boxplot-sex}
ggplot(abalone, aes(x=Sex, y=Age, fill=Sex)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=8, outlier.size=4) +
  scale_fill_manual(values=c(
    "F" = "pink",
    "I" = "lavender",
    "M" = "skyblue"
  )) +
  labs(
    title="Abalone Age by Sex",
    subtitle="Males show the highest median age, females the most outliers, and infants the youngest distribution",
    x="Sex",
    y="Age (years)"
  ) +
  theme_minimal()
```
```{r pairwise-plot}
library(GGally)

ggpairs(abalone, 
        columns = c("Length", "Diameter", "Height",
                    "Whole_weight", "Shucked_weight",
                    "Viscera_weight", "Shell_weight"),
        aes(color = Sex, alpha = 0.5)) +
  labs(
    title = "Pairwise Plot of Physical Measurements",
    subtitle = "Male and Female (Left skewed) and Infant (Symmetric), all are highly correlated"
  ) +
  theme_minimal()
```
**Observation:** Male and Female distributions are left skewed 
while Infant is symmetric. All physical measurements are highly 
correlated, indicating multicollinearity.
```{r descriptive-stats}
library(psych)

# Descriptive statistics for all numeric columns
describe(abalone[, c("Length", "Diameter", "Height",
                      "Whole_weight", "Shucked_weight",
                      "Viscera_weight", "Shell_weight", "Age")])
```
The descriptive statistics confirm the visual observations. Length and Diameter are negatively skewed reflecting that most abalones are fully grown, while Height is strongly positively skewed (3.13) indicating most abalones remain flat with few growing tall. Variables with similar mean and median values suggest symmetric distributions. Age shows the largest standard deviation, suggesting that abalones of similar size can vary considerably in age, highlighting the challenge of accurate age prediction.

H0 — "There is no difference in Age between Male, Female and Infant abalones"
H1 — "There is a significant difference in Age between at least one group"
```{r hypothesis-testing}
# Kruskal-Wallis test - comparing Age across Sex groups
result <- kruskal.test(Age ~ Sex, data = abalone)

# Clean output
cat("Kruskal-Wallis Test: Age by Sex\n")
cat("================================\n")
cat("Chi-squared:", round(result$statistic, 3), "\n")
cat("Degrees of freedom:", result$parameter, "\n")
cat("P-value:", format(result$p.value, scientific = TRUE), "\n")
cat("================================\n")
cat("Conclusion: We reject H0 - significant difference in Age between groups\n")
```
Based on the Kruskal–Wallis test, Sex has a statistically significant effect on Age.
```{r correlation}
# Correlation of all numeric variables with Age
cor(abalone[, c("Length", "Diameter", "Height",
                 "Whole_weight", "Shucked_weight",
                 "Viscera_weight", "Shell_weight", "Age")])
```
Shucked weight is the lowest because it represents the soft body meat which varies based on factors other than age such as nutrition and season.
```{r correlation-heatmap}
library(ggplot2)
library(reshape2)

# Correlation matrix
corr_matrix <- cor(abalone[, sapply(abalone, is.numeric)], use = "complete.obs")

# Convert to long format
corr_long <- melt(corr_matrix)

# Heatmap
ggplot(corr_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limit = c(-1, 1), name = "Correlation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    axis.title = element_blank()
  ) +
  coord_fixed() +
  labs(
    title = "Correlation Heatmap",
    subtitle = "Relationships Between Numerical Variables"
  )
```
Shell weight is the strongest predictor of Age, while even the weakest correlation (0.42) suggests all physical measurements carry some predictive value.